Spatio-temporal estimates of HIV risk group proportions for adolescent girls and young women across 13 priority countries in sub-Saharan Africa

### 2022-07-14 src/docs_aids22-poster/aids22-poster.Rmd
Poster draft out #111

---
main_topsize: 0.2
main_bottomsize: 0.075
poster_height: "48in"
poster_width: "36in"
font_family: "Open Sans"
# Essentials
title: Spatio-temporal estimates of HIV risk group proportions for adolescent girls and young women across 13 priority countries in sub-Saharan Africa
author:
  - name: '**Adam Howes**'
    affil: 1, 2
    main: true
    twitter: adamhowes
    email: ath19@ic.ac.uk
  - name: "Kathryn A. Risher"
    affil: 2, 3
  - name: "Van Kinh Nguyen"
    affil: 2
  - name: "Oliver Stevens"
    affil: 2
  - name: "Katherine M. Jia"
    affil: 4
  - name: "Tim M. Wolock"
    affil: 1, 2
  - name: "Lycias Zembe"
    affil: 5
  - name: "Ian Wanyeki"
    affil: 5
  - name: "Mary Mahy"
    affil: 5
  - name: "Clemens Benedikt"
    affil: 5
  - name: "Seth R. Flaxman"
    affil: 6
  - name: "Jeffrey W. Eaton"
    affil: 2
affiliation:
  - num: 1
    address: Department of Mathematics, Imperial College London
  - num: 2
    address: MRC Centre for Global Infectious Disease Analysis, School of Public Health, Imperial College London
  - num: 3
    address: Department of Medical Social Sciences, Feinberg School of Medicine, Northwestern University
  - num: 4
    address: Harvard T. H. Chan School of Public Health, Harvard University
  - num: 5
    address: Joint United Nations Programme on HIV/AIDS, Geneva, Switzerland
  - num: 6
    address: Department of Computer Science, University of Oxford
main_findings:
  - "Successful **targeting of HIV prevention** using **risk behaviour** would **greatly increase** expected new infections preemptively reached."
logoleft_name: "IMP_ML_W_N.png"
logoright_name: "UNAIDS-reverse-EN.gif"
# Style & formatting
column_numbers: 3
title_textsize: "60px"
author_textsize: "1.17em"
authorextra_textsize: "35px"
affiliation_textsize: "25px"
affiliation_textcol: '#00000080'
caption_fontsize: "20px"
# Main
main_fontfamily: "Open Sans"
# Poster body options
primary_colour: '#38761C'
secondary_colour: '#3b77D8'
accent_colour: "#5d19c4"
body_bgcol: "#ffffff"
body_textsize: "40px"
body_textcol: "#000000"
reference_textsize: "25px"
sectitle_textsize: "40px"
# Standard options
bibliography: citations.bib
output: 
  posterdown::posterdown_betterport:
    self_contained: false
    pandoc_args: --mathjax
    highlight: espresso
    number_sections: false
---

@@

* Adolescent girls and young women (AGYW) 15-29 **face disproportionately high risk** of HIV infection, and have been identified as a **priority population** for prevention efforts
* The UNAIDS Global AIDS Strategy 2021-2026 **recommends differentiating services** for AGYW geographically based both on individual risk behaviour and epidemic indicators
* We used a spatio-temporal model to estimate the proportion of AGYW in **four behavioural risk categories** $k = 1:4$ in 13 priority countries at a district level in the years 1999-2018

|$k$|Category                 | Risk ratio|
|:--|:------------------------|----------:|
|1  |Not sexually active      |          0|
|2  |One cohabiting partner   |          1|
|3  |Non-regular partner(s)   |       1.72|
|4  |Female sex workers (FSW) |         13|

@@

* We analyzed 47 national household surveys (AIS, BAIS, DHS, PHIA)
* For the categories $k \in \{1, 2, 3^{+}\}$ we fit a survey-weighted multinomial logistic regression via the multinomial-Poisson transformation
\begin{align}
    &y_{itak}^\star \sim \text{Poisson}(\kappa_{itak}), \\
    &\log(\kappa_{itak}) 
    = \underbrace{\theta_{ita}}_{\text{Observation (IID)}}
    + \underbrace{\beta_k}_{\text{Category (IID)}}
    + \underbrace{\alpha_{ak}}_{\text{Age (IID)}} \\
    &+ \underbrace{\zeta_{ck}}_{\text{Country (IID)}} 
    + \underbrace{\phi_{ik}}_{\text{Spatial (Besag)}} 
    + \underbrace{\gamma_{tk}}_{\text{Temporal (AR1)}}
\end{align}
where $p_{itak} = \text{softmax}(\kappa_{ita})_k$
* Using the Poisson formulation facilitated inference using integrated nested Laplace approximation [@rue2009approximate] via `R-INLA`
* To estimate the FSW proportion we used the 13 surveys with a transactional sex question to fit a survey-weighted logistic regression
\begin{align}
    &y_{ia4}^\star \sim \text{Binomial} \left( y_{ia3} + y_{ia4}, p_{ia4} / (p_{ia3} + p_{ia4}) \right), \\
    &\text{logit}(p_{ia4} / (p_{ia3} + p_{ia4}))
    = \beta_0
    + \beta_\texttt{cfswever} x_c \\
    &+ \underbrace{\alpha_a}_{\text{Age (IID)}}
    + \underbrace{\zeta_c}_{\text{Country (IID)}}
    + \underbrace{\phi_i}_{\text{Spatial (Besag)}}
\end{align}
where $x_c$ is the proportion of men in each country who are clients of FSW [@hodgins2021cfsw]
* We aligned our FSW estimates to the national-level UNAIDS Key Populations Atlas [@johnston2022deriving]
* We estimated incidence and number of expected new infections by risk group using the risk ratios to disaggregate Naomi model [@eaton2021naomi] general population incidence estimates

@@

```{r echo=FALSE, fig.align='center', out.width='100%', fig.cap='Dotplots showing the posterior mean of our district level estimates (in colour) and national estimates (in white) in 2018.'}
knitr::include_graphics("depends/3p1-within-between-country-variation.png")
```

```{r echo=FALSE, fig.align='center', out.width='100%', fig.cap='Cloropleths showing the posterior mean of our district level estimates in 2018.'}
knitr::include_graphics("depends/3p1-continental-map.png")
```

* There is **significant variation** in risk group proportions **within and between countries**, as well as **between age groups** (but close to no change over time)
* We found a geographic delineation, with **cohabiting more common in the east** and **non-regular partner(s) more common in the south**
* Large numbers of 15-19 in Mozambique have early sexual debut and are already cohabiting

```{r echo=FALSE, fig.align='center', out.width='100%', fig.cap='Comparison of targeting strategies'}
knitr::include_graphics("depends/infections-reached.png")
```

* Using location, age and behavioral risk stratification, **25% of expected new infections can be found by reaching 3% of the population**, compared with 8% of the population when behaviour is excluded 
* Majority of this benefit comes from reaching FSW, who are **3.5% of the at risk population but 21.0% of all expected new infections**

@@

* Using a **modular approach** allowed us to integrate all data from non-conformable surveys
* **Spatio-temporal smoothing** can be used to overcome high variance from the small sample sizes that result when further stratifying districts
* Individual behaviour is a **key determinant** of risk and providing prevention services on the basis of behavior would allow many more expected new infections to be reached, **especially among FSW**

@@

* No assessment of **practicalities or costs** of risk stratification approaches, or efficacy of available interventions
* No adjustment for **reporting bias**
* **Risk category definitions** could be disputed
  * What constitutes sex work?
  * Is within-group risk heterogeneity small enough?

@@

* Inclusion of **more surveys** e.g. VACS
* Extension to **general population**: men and women 15-49

>>>>

---
main_topsize: 0.2
main_bottomsize: 0.075
poster_height: "48in"
poster_width: "36in"
font_family: "Open Sans"
# Essentials
title: Spatio-temporal estimates of HIV risk group proportions for adolescent girls and young women across 13 priority countries in sub-Saharan Africa
author:
  - name: '**Adam Howes**'
    affil: 1, 2
    main: true
    twitter: adamhowes
    email: ath19@ic.ac.uk
  - name: "Kathryn A. Risher"
    affil: 2, 3
  - name: "Van Kinh Nguyen"
    affil: 2
  - name: "Oliver Stevens"
    affil: 2
  - name: "Katherine M. Jia"
    affil: 4
  - name: "Tim M. Wolock"
    affil: 1, 2
  - name: "Rachel Esra"
    affil: 2
  - name: "Lycias Zembe"
    affil: 5
  - name: "Ian Wanyeki"
    affil: 5
  - name: "Mary Mahy"
    affil: 5
  - name: "Clemens Benedikt"
    affil: 5
  - name: "Seth R. Flaxman"
    affil: 6
  - name: "Jeffrey W. Eaton"
    affil: 2
affiliation:
  - num: 1
    address: Department of Mathematics, Imperial College London
  - num: 2
    address: MRC Centre for Global Infectious Disease Analysis, School of Public Health, Imperial College London
  - num: 3
    address: Department of Medical Social Sciences, Feinberg School of Medicine, Northwestern University
  - num: 4
    address: Harvard T. H. Chan School of Public Health, Harvard University
  - num: 5
    address: Joint United Nations Programme on HIV/AIDS, Geneva, Switzerland
  - num: 6
    address: Department of Computer Science, University of Oxford
main_findings:
  - "Successful **priorisation of HIV prevention** using **risk behaviour** would **greatly increase** expected new infections preemptively reached."
logoleft_name: "IMP_ML_W_N.png"
# logoright_name: "UNAIDS-reverse-EN.gif"
# Style & formatting
column_numbers: 3
title_textsize: "60px"
author_textsize: "1.17em"
authorextra_textsize: "35px"
affiliation_textsize: "25px"
affiliation_textcol: '#00000080'
caption_fontsize: "20px"
# Main
main_fontfamily: "Open Sans"
# Poster body options
primary_colour: '#38761C'
secondary_colour: '#3b77D8'
accent_colour: "#5d19c4"
body_bgcol: "#ffffff"
body_textsize: "40px"
body_textcol: "#000000"
reference_textsize: "25px"
sectitle_textsize: "40px"
# Standard options
bibliography: citations.bib
output: 
  posterdown::posterdown_betterport:
    self_contained: false
    pandoc_args: --mathjax
    highlight: espresso
    number_sections: false
---

@@

* Adolescent girls and young women (AGYW) 15-29 face disproportionately high risk of HIV infection, and are a priority population for prevention efforts
* The Global AIDS Strategy 2021-2026 recommends differentiating services for AGYW geographically based both on individual risk behaviour and epidemic indicators
* We used a spatio-temporal model to estimate the proportion of AGYW in four behavioural risk categories in 13 priority countries at a district level in the years 1999-2018

|$k$|Risk group               | Risk ratio|
|:--|:------------------------|----------:|
|1  |Not sexually active      |          0|
|2  |One cohabiting partner   |          1|
|3  |Non-regular or multiple partner(s)   |       1.72|
|4  |Female sex workers (FSW) |         13|

@@

* We analyzed 46 national household surveys, including AIS, BAIS, DHS and PHIA
* We took a two-stage modelling approach:
1. We fit a survey-weighted multinomial logistic regression to estimate the proportion of AGYW each risk group, with the non-regular or multiple partner(s) and FSW risk groups combined together, using age (IID), country (IID), spatial (ICAR) and temporal (IID) random effects
2. To estimate the FSW proportion we used the 13 surveys with a transactional sex question to fit a survey-weighted logistic regression using age (IID), country (IID) and spatial (ICAR) random effects, as well as covariates for the proportion of men who are clients of FSW [@hodgins2022population]
* We aligned our FSW estimates to national-level estimates obtained by disaggregating @stevens2022estimating by age group
* We disaggregated Naomi model [@eaton2021naomi] general population incidence and prevalence estimates by risk group using incidence risk ratios and prevalence ratios
* We estimated the new infections that could be reached prioritisting according to different strategies

\newpage

@@

```{r echo=FALSE, fig.align='center', out.width='100%', fig.cap='District level estimates (in colour) and national estimates (in white) in 2018.'}
knitr::include_graphics("depends/3p1-within-between-country-variation.png")
```

```{r echo=FALSE, fig.align='center', out.width='100%', fig.cap='Spatial distribution of our 2018 estimates.'}
knitr::include_graphics("depends/3p1-continental-map.png")
```

* Risk group proportions varied substantially across age groups (65.9% of total variation explained), countries (20.9%), and between districts within each country (11.3%), but changed little over time (0.9%)
* Among women aged 20-29, cohabiting (63.1%) was more common in eastern Africa than non-regular or multiple partner(s) (21.4%), while in southern countries non-regular or multiple partner(s) (58.5%) were more common than cohabiting (23.4%)
* Large numbers of 15-19 in Mozambique have early sexual debut (64.2%) and close to a third (34.2%) are already cohabiting

```{r echo=FALSE, fig.align='center', out.width='100%', fig.cap='Comparison of prioritisation strategies'}
knitr::include_graphics("depends/infections-reached.png")
```

* Using location, age and behavioral risk prioritisation, 50% of expected new infections can be found by reaching 10.6% of the population, compared with 19.3% of the population when behaviour is excluded 
* The majority of this benefit comes from reaching FSW, who are 1.4% of the at risk population but 10.9% of all expected new infections

@@

* Providing prevention services on the basis of behavior would allow many more expected new infections to be reached, especially among FSW
* Programs should ensure that behavioural prioritisation occurs without stigmatising or blaming AGYW
* We focused on the most proximal determinants of risk, based on location and behaviour, rather than more distal determinants
* Using a two-stage approach allowed us to include all surveys, even those without a specific transactional sex question
* We overcome the small sample sizes created by multiple stratifications by using spatio-temporal smoothing

@@

* No adjustment for under-reporting biases, which likely vary by age group
* Choosing appropriate risk group definitions that succinctly capture variation in risk consistently across contexts is challenging
* We are least confident in our FSW estimates, which face particular measurement difficulties
* We did not assess the practicalities or costs of the risk prioritisation strategies we considered

@@

* Inclusion of more surveys e.g. VACS
* Extension to general population (adults 15-49)

<font size="6"> **Interested to read more?** Manuscript and R code available are available from `github.com/athowes/multi-agyw`. </font>
